---
title: <center><strong> Web scraping using R </strong></center>
author: <center><strong> Christopher Maronga </strong></center>
categories: databases with R
output: 
  html_document:
    number_sections: no
    theme: 
      bootswatch: minty
      base_font: "sans-serif"
      primary: "#008B8B"
      secondary: "#002147"
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: yes
      smooth_scroll: yes
---

<style>
h1, h2, h3, h4, h5, h6, legend {
    color: #008B8B;
}

#nav-top span.glyphicon {
    color: #22983B;
}

#table-of-contents header {
    color: #22983B;
}

#table-of-contents h2 {
    background-color: #22983B;
}

#main a {
    background-image: linear-gradient(180deg,#d64a70,#d64a70);
    color: #c7254e;
}

a:hover {
    color: #ff0000;
}

  #content {
    max-width: 13000px;
  }
  
  #sidebar h2 {
    background-color: #008B8B;
  }

</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, 
                      message = F,
                      echo = T, 
                      dpi = 180, 
                      fig.width = 12, 
                      fig.height = 8,
                      comment = "")
library(ggpubr)
library(ggthemes)
library(bslib)
# some house rules for plots and tables
theme_set(theme_economist(base_size = 20))
```


## Introduction

Web scraping is concept that most probably you have heard about. This is the art of [harvesting](https://en.wikipedia.org/wiki/Web_scraping) publicly available data from a website for use in your analysis or reporting. Web scraping can be as simple as copying the contents of a website and pasting them on an excel sheet, but that's not what we are going to do today. Most website pages are built using [`HTML`](https://en.wikipedia.org/wiki/HTML) and this allows us to use tools such as R to dynamically extract the data.

In this tutorial, I am going walk you through how you can harvest data from websites using R programming language. You can do this by coding the logic and instructions manually or use the package `rvest` to easily extract the contents of a website. We would demonstrate the examples using the below two websites

- National Institute for Health and care Excellence ([NICE](https://bnf.nice.org.uk/drug/)) - List of drugs and pricing
- The Global Economy.com [Dollar](https://www.theglobaleconomy.com/rankings/Dollar_exchange_rate/) exchange rates for various countries



## Loading packages

```{r packages}
library(tidyverse) # data wrangling and viz
library(rvest)     # for webscraping
library(ggthemes)
library(foreach)
library(DT)
```


## Webscraping BNF website

The British National Formulary ([BNF](https://bnf.nice.org.uk/drug/)) and British National Formulary for Children ([BNFC](https://bnfc.nice.org.uk/drug/)) websites contains  up to date and highly authoritative information resource on medicines prescribed in the UK. These resources are mostly used by doctors, pharmacists and other healthcare professionals and it aims to provide healthcare professionals with sound up-to-date information about the use of and cost of medicines


Typically, healthcare professionals using information from this website need to access it by searching through the alphabets for the drug of interest, copy & paste the information they are looking for. This proves to be quite tedious and time-consuming more especially if you need to do this for a long list of drugs.
In this article, I am going to show you how you can dynamically fetch this information using R a concept known as web scraping or web-harvesting.


First, you need to navigate to the page that contains the drug information that you want to extract. In this tutorial, I am going to use [paracetamol](https://bnf.nice.org.uk/medicinal-forms/paracetamol.html); but note that the below approach can be used with any drug in the website.

Use the `read_html` to read the HTML code from the website url

```{r}
# get url from website
URL <- "https://bnf.nice.org.uk/medicinal-forms/paracetamol.html"

# read HTML from website
paracetamol_webpage <- read_html(URL)

# structure of `paracetamol_webpage`
glimpse(paracetamol_webpage)

```


### Extract nodes

The contents of the object `paracetamol_webpage` contains xml nodes which are essentially the number of tables from the webpage. we extract this information using the function `html_nodes()` that takes the html webpage object `paracetamol_webpage` as it's first argument and `table` as the second argument as shown below.

```{r}
# extract nodes (tables)
paracetamol_nodes <- html_nodes(paracetamol_webpage, 
                               "table") # extract all nodes (containing all tables)


length(paracetamol_nodes); head(paracetamol_nodes)
```


This simply means that the paracetamol webpage contains 167 tables, and we are going to fetch all these tables and combine them into one long list/data frame. The object `paracetamol_nodes` contains the 167 tables that can be accessed using the function`html_table` and the object name as it's first argument.

```{r}
# use html_table to extract tables from the nodes object
paracetamol_tables <- html_table(paracetamol_nodes)

# list the first two tables
head(paracetamol_tables, 2)

```

`paracetamol_tables` is a complex list (I unofficially call it `nested list`) that contains the table information as the first index of each list entry i.e. out of the 167 elements of the object, which are all lists in themselves, extracting position of each of them returns a nice tibble which is the table that we seek to extract.

For instance, if I want table number 3 from the object `paracetamol_tables`, I will write the following code

```{r}
# extract table 3 of 167
paracetamol_tables[3][[1]]
```


```{r}
# extract table 8 of 167
paracetamol_tables[8][[1]]
```


Take **`NOTE`** of the indices that change based on the number of tables returned by `html_nodes` and the use of double `[[]]` for the second index.


### Extracting the tables

We will write a loop using `foreach` package to automatically extract the individuals tables and bind them as tibble that can used to easily filter and search for information of interest in the columns.

```{r}
# pass the table IDs a
paracetamol_df <- foreach(tableID = seq_along(paracetamol_tables), .combine = bind_rows)%do%{
  # rbind to form tibble
  paracetamol_tables[tableID][[1]]
}

tail(paracetamol_df)
```

Fantastic, job well done. We have extracted all the tables from the paracetamol webpage and combined them into a tibble that can be exported into a CSV. Our first task for this web scraping is now done. You can try replicate the above flow by picking a different drug and following through. The table below now shows the complete list of the paracetamol medication information that we just harvested from the BNF website.

```{r}
paracetamol_df %>% datatable(rownames = F)
```


## Dollar exchange rates

The previous example using the BNF website was pretty much straight forward and easy, however not all websites will be well structured as the BNF website and thus, the `rvest` functions we saw above will fail, hence the above approach won't work. Therefore, this calls for a manual web scraping approach without relying on `rvest` package. A good example is the next [website](https://www.theglobaleconomy.com/rankings/Dollar_exchange_rate/) on dollar exchange rates for various countries.

















































