---
title: <center><strong> Web scraping using R </strong></center>
author: <center><strong> Christopher Maronga </strong></center>
categories: databases with R
output: 
  html_document:
    number_sections: no
    code_folding: hide
    theme: united
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: yes
      smooth_scroll: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, message = F,
                      echo = T, 
                      dpi = 180, 
                      fig.width = 12, 
                      fig.height = 8,
                      results = "asis")
```


# Introduction

Web scraping is concept that most probably you have heard about. This is the art of [harvesting](https://en.wikipedia.org/wiki/Web_scraping) publicly available data from a website for use in your analysis or reporting. Web scraping can be as simple as copying the contents of a website and pasting them on an excel sheet, but that's not what we are going to do today. Most website pages are built using [`HTML`](https://en.wikipedia.org/wiki/HTML) and this allows us to use tools such as R to dynamically extract the data.

In this tutorial, I am going walk you through how you can harvest data from websites using R programming language. You can do this by coding the logic and instructions manually or use the package `rvest` to easily extract the contents of a website. We would demonstrate the examples using the below two websites

- National Institute for Health and care Excellence ([NICE](https://bnf.nice.org.uk/drug/)) - List of drugs and pricing
- The Global Economy.com [Dollar](https://www.theglobaleconomy.com/rankings/Dollar_exchange_rate/) exchange rates for various countries



# Loading packages

```{r packages}
library(tidyverse) # data wrangling and viz
library(rvest)     # for webscraping
```












